---
title: "Mission Notes and Submission Logging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mission Notes and Submission Logging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Mission Notes System

The BIOsubmissions package includes a comprehensive logging system to track submission notes, troubleshooting steps, and manual adjustments for each mission. This system helps maintain institutional knowledge and ensures consistency across submissions.

## Overview

The mission notes system provides:

- **Organized storage** of notes by platform (OCADS, SDG) and mission
- **Easy retrieval** of historical submission information
- **Search functionality** to find specific issues or solutions
- **Interactive prompts** to capture notes after each submission
- **Structured format** for consistency

## Directory Structure

Notes are stored in the `log/` directory organized by platform and mission:

```
log/
├── OCADS/
│   ├── BBMP/
│   │   ├── 2021.md
│   │   └── 2022.md
│   ├── AZOMP/
│   │   ├── AT4802_2022.md
│   │   └── AT4805_2022.md
│   └── JC/
│       └── JC24301.md
└── SDG/
    └── 2024.md
```

## Basic Usage

### Reading Notes

To read notes for a specific mission:

```{r, eval=FALSE}
library(BIOsubmissions)

# Read BBMP 2022 notes
notes <- read_mission_notes("OCADS", "BBMP", 2022)
cat(notes, sep = "\n")

# Read JC24301 notes
notes <- read_mission_notes("OCADS", "JC24301")
cat(notes, sep = "\n")
```

### Adding Notes

#### After a Submission (Interactive)

The easiest way to add notes is using the interactive prompt after completing a submission:

```{r, eval=FALSE}

# Prompt for notes
prompt_submission_notes("OCADS", "BBMP", 2023)
```

This will:
1. Show existing notes (if any)
2. Ask if you want to add notes
3. Let you type notes line by line
4. Automatically timestamp and format your entry

#### Programmatically Adding Notes

You can also add notes programmatically:

```{r, eval=FALSE}
# Append to existing notes
append_mission_notes(
  platform = "OCADS",
  mission = "BBMP",
  year = 2023,
  notes = c(
    "Fixed pH metadata issue",
    "Updated investigator list",
    "Accession: 0228686"
  )
)
```

### Listing All Missions

To see all missions with notes:

```{r, eval=FALSE}
# List all missions
all_missions <- list_mission_notes()
print(all_missions)

# List only OCADS missions
ocads_missions <- list_mission_notes("OCADS")
print(ocads_missions)
```

### Searching Notes

Search across all mission notes for specific information:

```{r, eval=FALSE}
# Search for pH-related issues
ph_issues <- search_mission_notes("pH")
print(ph_issues)

# Search for accession numbers in OCADS submissions
accessions <- search_mission_notes("accession", platform = "OCADS")
print(accessions)

# Case-sensitive search
exact_match <- search_mission_notes("CTD", ignore_case = FALSE)
```

## Automatic Integration

**The conversion functions now automatically prompt for notes!**

When you use `convert_OCADS()` or `convert_CCHDO()`, the functions will automatically prompt you to add submission notes after the conversion is complete.

### OCADS Conversion with Auto-Prompting

```{r, eval=FALSE}
# Default behavior: prompts for notes after conversion
ocads_data <- convert_OCADS(bcd_data, biochem.password, biochem.username)
# After conversion completes, you'll be prompted:
# "Would you like to add submission notes? (y/n):"

# To disable prompting (if you want to add notes later)
ocads_data <- convert_OCADS(bcd_data, biochem.password, biochem.username, 
                            prompt_notes = FALSE)
```

### CCHDO Conversion with Optional Prompting

```{r, eval=FALSE}
# Default: no prompt (notes typically added during OCADS conversion)
cchdo_data <- convert_CCHDO(ocads_data)

# Enable prompting for CCHDO-specific submissions
cchdo_data <- convert_CCHDO(ocads_data, prompt_notes = TRUE)
```

### Benefits of Auto-Prompting

- **Don't forget**: Captures notes right after conversion while details are fresh
- **Extract info automatically**: Mission and year information extracted from data
- **Optional**: Can disable if you prefer to add notes manually later
- **Graceful fallback**: If prompting fails, you can add notes later

### Example Full Workflow

```{r, eval=FALSE}
library(BIOsubmissions)
library(tidyverse)

# Read your data
bcd_data <- read_csv("mission_BCD.csv")

# Convert to OCADS (will prompt for notes)
ocads_data <- convert_OCADS(bcd_data, biochem.password, biochem.username)
# ✓ OCADS conversion complete!
# Mission: 18HU23573 (EXPOCODE: 18HU20230504)
# Would you like to add submission notes? (y/n): y
# [You type your notes...]

# Convert to CCHDO (no prompt by default)
cchdo_data <- convert_CCHDO(ocads_data)

# Save your files
write_csv(ocads_data, "mission_OCADS.csv")
write_csv(cchdo_data, "mission_CCHDO.csv", quote = 'none')
```

## Typical Workflow

### 1. Before Submission

Check for previous notes to learn from past issues:

```{r, eval=FALSE}
# Check what issues were encountered before
notes <- read_mission_notes("OCADS", "BBMP", 2022)
cat(notes, sep = "\n")

# Search for specific known issues
flag_issues <- search_mission_notes("flag")
```

### 2. During Submission

Keep track of any manual adjustments or issues encountered.

### 3. After Submission

Document everything you did:

```{r, eval=FALSE}
# Interactive prompt
prompt_submission_notes("OCADS", "BBMP", 2023)

# Or programmatically
append_mission_notes(
  platform = "OCADS",
  mission = "BBMP", 
  year = 2023,
  notes = c(
    "## Submission Summary",
    "- Accession: 0228686",
    "- Submitted to: alex.kozyr@noaa.gov",
    "- Date: 2023-05-15",
    "",
    "## Issues Encountered",
    "- Had to manually fix date formatting",
    "- Updated CTDFLUOR sampling instrument to CTD",
    "",
    "## Action Items",
    "- [ ] Follow up on pH metadata with lab",
    "- [ ] Verify with Diana on funding numbers"
  )
)
```

## What to Document

Good notes should include:

### Essential Information
- **Accession number**
- **Submission date**
- **Contact person/email**
- **Submission status** (submitted, pending, resubmitted)

### Technical Details
- **Data issues** encountered and how they were resolved
- **Metadata corrections** made manually
- **Script bugs** discovered
- **Manual adjustments** to files

### Follow-up Items
- **Action items** for future submissions
- **Questions** to ask colleagues
- **Verification tasks** needed

### Lessons Learned
- **What worked well**
- **What to avoid** next time
- **Useful patterns** or solutions

## Example Entry

Here's an example of a well-documented submission:

```markdown
# OCADS - BBMP 2023

**Created:** 2023-05-15
**Last Updated:** 2023-05-15
**Accession Number:** 0240502

---

## Submission Details

- **Status:** Submitted
- **Submission Date:** 2023-05-15
- **Contact:** alex.kozyr@noaa.gov
- **Method:** Excel submission

## Issues Encountered

### Date Formatting
- **Problem:** Dates from script formatted as YYYYMMDD instead of ISO format
- **Solution:** Manually reformatted to YYYY-MM-DD
- **Future:** Need to fix in data processing script

### Metadata Corrections
- Updated investigator names
- Fixed special characters in abstract
- Changed CTDFLUOR to CTD in sampling instrument

## Files Submitted
- `BBMP_2023_data.csv`
- `BBMP_2023_metadata.xlsx`
- Updated mission list

## Action Items
- [ ] Fix date formatting in processing script
- [ ] Create special character cleaning function
- [ ] Update instrument mapping in metadata template

---

*Submitted by: E. O'Grady*
```

## Advanced Usage

### Creating New Mission Notes

To start notes for a new mission:

```{r, eval=FALSE}
write_mission_notes(
  platform = "OCADS",
  mission = "BBMP",
  year = 2024,
  notes = c(
    "# OCADS - BBMP 2024",
    "",
    "**Created:** " %+% Sys.Date(),
    "",
    "---",
    "",
    "## Submission Notes",
    ""
  )
)
```

### Batch Operations

Process notes across multiple missions:

```{r, eval=FALSE}
# Get all OCADS missions
missions <- list_mission_notes("OCADS")

# Check each for specific issue
for (i in seq_len(nrow(missions))) {
  notes <- readLines(missions$file[i])
  if (any(grepl("pH", notes, ignore.case = TRUE))) {
    cat("pH mentioned in:", missions$mission[i], "\n")
  }
}
```

## Tips and Best Practices

1. **Document immediately** - Add notes right after submission while details are fresh

2. **Be specific** - Include exact error messages, file names, and values changed

3. **Use consistent formatting** - Follow the markdown structure in examples

4. **Include context** - Explain *why* you made changes, not just *what* changed

5. **Update regularly** - If you learn something new about an old submission, add a note

6. **Search before asking** - Use `search_mission_notes()` to find previous solutions

7. **Cross-reference** - Link related issues across missions

8. **Action items** - Use `- [ ]` for unchecked tasks, `- [x]` for completed ones



## Troubleshooting

### Notes file not found

If you get "No notes found", the file may not exist yet:

```{r, eval=FALSE}
# Create initial notes
append_mission_notes("OCADS", "NewMission", 2024, 
                     notes = "Initial submission")
```

### Search returns no results

Try broader search terms or check spelling:

```{r, eval=FALSE}
# Too specific
search_mission_notes("pH instrument calibration")

# Better
search_mission_notes("pH")
```

## See Also

- `?read_mission_notes` - Full function documentation
- `?append_mission_notes` - Adding notes documentation
- `?search_mission_notes` - Search functionality
- [Submission Guide](submission-guide.html) - Complete submission workflow
