---
title: "Quick Start: Processing Oceanographic Data for International Submissions"
author: "BIO Data Management Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start: Processing Oceanographic Data for International Submissions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This guide will walk you through processing oceanographic data from BioChem and preparing it for submission to international data platforms (OCADS, CCHDO, SDG). The workflow is designed to be reproducible and catch common data quality issues that have historically caused submission rejections.

## What This Package Does

The `BIOsubmissions` package:

- Converts BioChem BCD (BioChem Data) format to international submission formats
- Performs unit conversions (e.g., mL/L to µmol/kg for oxygen)
- Translates BioChem method names to standard parameter names (CCHDO/OCADS)
- Applies quality flags according to WOCE standards
- Validates data ranges for latitude, longitude, and oceanographic parameters

## Prerequisites

Before you begin, you'll need:

1. **BioChem Access**: Username and password for the BioChem database
2. **R Environment**: R (>= 4.0.0) with RStudio recommended
3. **Required Packages**: tidyverse, ROracle, DBI, RSQLite, oce

# Installation

Install the package from GitHub:

```{r install}
# Install devtools if needed
if (!require("devtools")) install.packages("devtools")

# Install BIOsubmissions
devtools::install_github("eogrady21/BIOsubmissions")
```

Load the package and dependencies:

```{r setup}
library(BIOsubmissions)
library(tidyverse)
library(here)
```

# Step-by-Step Workflow

## Step 1: Set Up BioChem Credentials

Create a file to store your BioChem credentials (NEVER commit this to Git!):

```{r credentials}
# Create a file: C:/users/YOUR_USERNAME/desktop/biochem_creds.R
# Contents:
biochem.user <- "your_username"
biochem.password <- "your_password"
```

Add this file to your `.gitignore`:

```
biochem_creds.R
*.biochem_creds.R
```

## Step 2: Extract Data Directly from BioChem

**NEW:** You can now extract data directly from BioChem without using SQL Developer!

```{r extract-biochem}
# Source your credentials
source("C:/users/ogradye/desktop/biochem_creds.R")

# Extract data directly from BioChem for your mission
data <- extract_from_biochem(
  mission_descriptor = "LAT2025146",
  biochem.username = biochem.user,
  biochem.password = biochem.password
)

# Check the structure
glimpse(data)
```

**What `extract_from_biochem()` does:**
- Connects to BioChem automatically
- Executes the BCD query for your mission
- Returns data in proper BCD format with all required columns
- Formats dates correctly (%m/%d/%Y)
- Adds standard BioChem metadata columns
- Performs preliminary validation:
  - Checks for core parameters (Temperature, Salinity, Pressure)
  - Identifies carbonate chemistry and tracer parameters
  - Flags missing or suspicious QC codes
  - Validates coordinates

**Optional: List Available Missions**

Not sure of the exact mission descriptor? You can query BioChem:

```{r list-missions}
# List all missions from 2025
missions_2025 <- list_biochem_missions(
  biochem.username = biochem.user,
  biochem.password = biochem.password,
  year = 2025
)

# List missions for a specific ship
missions_atlantis <- list_biochem_missions(
  biochem.username = biochem.user,
  biochem.password = biochem.password,
  ship_code = "18AT"
)
```

### Alternative: Manual SQL Export (Legacy Method)

If you prefer the traditional workflow or need to work offline:

1. Open `inst/BCD_QUERY.sql` in SQL Developer
2. Update the mission descriptor in the WHERE clause
3. Run the query and export to CSV
4. Read the CSV in R:

```{r load-csv}
data <- read_csv("LAT2025146_BCD.csv", show_col_types = FALSE)
```

## Step 3: Prepare Your Data (if needed)

### Data Preparation Notes

**If using `extract_from_biochem()`:** Data is already properly formatted! Skip to Step 4.

**If your BCD data was extracted via SQL Developer**, you may need to manually fix some columns:

```{r fix-bcd}
# Fix date format if needed (SQL Developer may export as 'DD-MON-YY')
data$DIS_HEADER_SDATE <- format(
  as.Date(data$DIS_HEADER_SDATE, format = '%d-%b-%y'),
  format = '%m/%d/%Y'
)

# Add required columns if missing
if (!"DIS_DATA_NUM" %in% names(data)) {
  data$DIS_DATA_NUM <- seq(1:nrow(data))
}

if (!"CREATED_BY" %in% names(data)) {
  data$CREATED_BY <- Sys.getenv("USERNAME")  # Your Windows username
}

if (!"CREATED_DATE" %in% names(data)) {
  data$CREATED_DATE <- Sys.Date()
}

# Add standard BioChem fields
data$DATA_CENTER_CODE <- '20'  # BIO data center code
data$PROCESS_FLAG <- 0
data$BATCH_SEQ <- 1
```

### Handling Multiple Data Types for Same Parameter

If you have both fresh and frozen nutrient samples, combine them to avoid naming conflicts:

```{r consolidate-methods}
data <- data %>%
  mutate(DATA_TYPE_METHOD = case_when(
    DATA_TYPE_METHOD %in% c('NH3_Filt_Fsh', 'NH3_Filt_F') ~ 'NH3_0',
    DATA_TYPE_METHOD %in% c('NO2_Filt_Fsh', 'NO2_Filt_F') ~ 'NO2_0',
    DATA_TYPE_METHOD %in% c('NO2NO3_Filt_Fsh', 'NO2NO3_Filt_F') ~ 'NO2NO3_0',
    DATA_TYPE_METHOD %in% c('PO4_Filt_Fsh', 'PO4_Filt_F') ~ 'PO4_0',
    DATA_TYPE_METHOD %in% c('SiO4_Filt_Fsh', 'SiO4_Filt_F') ~ 'SiO4_0',
    TRUE ~ DATA_TYPE_METHOD
  ))
```

## Step 4: Convert to OCADS Format

The `convert_OCADS()` function handles all the complex transformations:

```{r convert-ocads}
# Convert to OCADS format
ocads_data <- convert_OCADS(data, biochem.password, biochem.user)

# Save the output
write_csv(ocads_data,
          file = here("data", "2025", "LAT2025146", "LAT2025146_data.csv"))
```

**What this function does:**
- Connects to BioChem to retrieve sounding (bottom depth) data
- **Validates data completeness**: Checks for carbonate chemistry parameters (ALKALI, PH_TOT, TCARBN, PCO2)
- **Validates tracer data**: Checks for CFC-12, SF6, and DELO18 parameters
- **Validates QC flags**: Identifies missing or suspicious flags (all zeros indicate no QC applied)
- Translates BioChem method names to CCHDO/OCADS parameter names using lookup tables
- Converts units (oxygen: mL/L → µmol/kg, nutrients: µmol/L → µmol/kg)
- Translates BioChem QC flags (0-9) to WOCE flags (2,3,4,6,9)
- Averages replicate measurements and assigns flag 6
- Validates data integrity with built-in spot checks

**Data Source:** All data values and QC flags are extracted directly from BioChem. The function
will warn you if expected parameters are missing or if QC flags haven't been properly applied
in BioChem.

## Step 5: Convert to CCHDO Format (if needed)

CCHDO doesn't accept an ordinary CSV - it requires the WHP-exchange bottle format: a
creation-stamp line, a units row, and an `END_DATA` terminator, in addition to the usual
header/data rows (see https://exchange-format.readthedocs.io/). `convert_CCHDO()` reshapes
the OCADS output to match; `write_CCHDO_exchange()` writes the actual `<EXPOCODE>_hy1.csv`
file - don't `write_csv()` the result of `convert_CCHDO()` directly, it's missing that
structure.

```{r convert-cchdo}
# Convert OCADS output to CCHDO/WHP-exchange format.
# Defaults check each station's position against the AR07W line (station 1 to station 17)
# and only label it "AR07W" if it's within 50 km of that line - AZOMP also occupies
# stations that aren't on AR07W (Scotian Slope/Rise, Bedford Basin, etc.), and those are
# left with a blank SECT_ID rather than mislabelled. Review the "[SECT_ID]" console
# messages and fill in blanks manually if they belong to a different, known section.
cchdo_data <- convert_CCHDO(ocads_data)

# Writes LAT2025146_hy1.csv (the required WHP-exchange bottle filename)
write_CCHDO_exchange(
  cchdo_data,
  out_dir = here("data", "2025", "LAT2025146", "CCHDO")
)
```

**Key CCHDO/WHP-exchange differences:**
- File structure: stamp line, units row, and `END_DATA` terminator (not a plain CSV)
- File name: `<EXPOCODE>_hy1.csv`, not a mission-descriptor name
- DATE format: YYYYMMDD (no dashes)
- Column names: LATITUDE/LONGITUDE instead of BTL_LAT/BTL_LON; SECT_ID added, but only for
  stations that geographically fall on the line (see below) - check the console output
- DEPTH: WHP `DEPTH` is bottom depth (OCADS's `SOUNDING`), not the per-bottle sample depth
  OCADS calls `DEPTH` - that column is dropped since it was already used to derive `CTDPRS`
- Station names: alphanumeric only
- NH3 renamed to NH4

## Step 6: Convert to SDG 14.3.1 Format (if needed)

The SDG 14.3.1 indicator tracks ocean acidification and requires a specific Excel (`.xlsx`) format.
The `convert_SDG()` function takes your OCADS output and produces a submission-ready file with
no manual editing required.

```{r convert-sdg}
# Convert OCADS output to SDG 14.3.1 format
# The function writes the .xlsx file and returns the data invisibly
sdg_data <- convert_SDG(
  OCADS_fn = here("data", "2025", "LAT2025146", "LAT2025146_data.csv")
)
```

The output file will be written automatically to:

```
data/SDG/<MOORING_NAME>/<MOORING_NAME>_data_SDG.xlsx
```

### SDG Platform Requirements

The SDG 14.3.1 data portal enforces strict formatting rules. `convert_SDG()` satisfies all of
them automatically:

| Requirement | Column | How it is handled |
|---|---|---|
| Date as `YYYY-MM-DD` | `DATE_UTC` | Normalised through `as.Date()` regardless of input format |
| Time as `HH:MM:SS` | `TIME_UTC` | Padded / reformatted to always produce 8-character strings |
| Coordinates as Number | `LONGITUDE`, `LATITUDE` | Coerced with `as.numeric()` |
| All variable columns as Number | all non-metadata cols | `mutate(across(..., as.numeric))` applied before writing |
| `.xlsx` file format | — | Written with `openxlsx`; cell types set explicitly |

> **Why `openxlsx` matters:** The older `xlsx` package writes all values as character strings
> regardless of the R class. `openxlsx` sets proper Excel cell types (Number, Text) so the SDG
> portal recognises numeric columns without any manual reformatting in Excel.

### What `convert_SDG()` Does

- Loads the SDG template (`SDG_14_3_1-data_submission-2019.xlsx`) to preserve required header
  structure
- Maps OCADS column names to SDG names using the `ocads_sdg_data.xlsx` lookup table
- Renames flag columns from `_FLAG_W` to ` FLAG` convention
- Derives `DEPTH_SAMPLING` from CTD pressure (via `oce::swDepth()`) if the depth column is
  missing
- Applies `YYYY-MM-DD` date style and `HH:MM:SS` time style to the correct Excel columns
- Applies a numeric cell format to all variable columns

### Troubleshooting SDG Submissions

**"The date column is not recognised"**

The portal expects `YYYY-MM-DD`. Check your input file for slash-delimited dates
(`MM/DD/YYYY`). `convert_SDG()` will warn you and correct this automatically, but verify the
output if you see:

```
Warning: Slash-delimited date format detected — converting to YYYY-MM-DD.
```

**"Coordinates are flagged as text"**

This can happen if the upstream OCADS file has coordinates stored as character strings (e.g.,
`"44.6700N"`). Ensure coordinates in your OCADS file are plain decimal degree numbers with no
hemisphere suffix before calling `convert_SDG()`.

**"TIME_UTC is all NA — falling back to BTL_TIME"**

Some OCADS exports store bottle time in `BTL_TIME` rather than `TIME_UTC`. The function falls
back automatically, but review the output to confirm times are populated correctly.

**Depth column missing**

If `DEPTH_SAMPLING` is empty in your OCADS file, `convert_SDG()` will derive it from
`CTDPRS` using the `oce::swDepth()` function and issue a warning:

```
Warning: DEPTH_SAMPLING contains NAs — deriving depth from CTDPRS.
```

Verify the derived depths are reasonable before submission.

### Reference Files Required

`convert_SDG()` expects two reference files in your working directory:

| File | Purpose |
|---|---|
| `SDG_14_3_1-data_submission-2019.xlsx` | Official SDG template (provides header rows) |
| `ocads_sdg_data.xlsx` | Column name mapping between OCADS and SDG |

Contact the data manager if either file is missing or needs to be updated with new parameters.

# Understanding the Output

## Data Validation Warnings

During conversion, you may see validation messages about data completeness:

### Carbonate Chemistry Parameters

```
[NOTE] Missing carbonate chemistry parameters: ALKALI, PH_TOT
```

**What it means:** The expected carbonate parameters weren't found in BioChem for this mission.

**Action required:**
- If this mission collected carbonate chemistry data, ensure it has been uploaded to BioChem with proper QC flags
- If this mission didn't collect carbonate data, this warning is expected and can be ignored

### Tracer Parameters

```
[NOTE] Missing tracer parameters: CFC-12, SF6, DELO18
```

**What it means:** The expected tracer gases weren't found in BioChem for this mission.

**Action required:**
- If this mission collected tracer data, ensure it has been uploaded to BioChem
- If this mission didn't collect tracer data, this warning is expected and can be ignored

### QC Flag Warnings

```
[!] All QC flags for ALKALI are 0 - no quality control has been applied!
```

**What it means:** All flags are zero, suggesting that data was uploaded to BioChem without quality control.

**Action required:**
- Return to BioChem and apply proper QC flags (1-9 scale) before submission
- Review lab QC procedures and CRM analyses
- Update flags in BioChem, then re-extract and re-run conversion

```
[!] TCARBN has data values but missing QC flags!
```

**What it means:** Data exists but no quality flags were found.

**Action required:**
- Add QC flags to BioChem for this parameter
- Flags should be based on CRM accuracy, replicate precision, and analyst review

## File Structure

The output CSV files have a special structure:

```
Row 1: Units (e.g., METERS, µmol/kg, PSU)
Row 2+: Data values
```

Example:
```
EXPOCODE,STNNBR,DEPTH,CTDSAL,NO2+NO3,NO2+NO3_FLAG_W
18QL20230504,AR7W01,METERS,PSU,µmol/kg,
18QL20230504,AR7W01,5.2,35.1,12.3,2
```

## Quality Flags (WOCE Standards)

The `_FLAG_W` columns use WOCE quality codes:

- **2**: Acceptable (good quality)
- **3**: Questionable (use with caution)
- **4**: Bad (do not use)
- **6**: Replicate measurement (average of multiple samples)
- **9**: Missing data (value = -999)

## Missing Data Convention

Missing or unavailable data is represented as:
- Data values: `-999`
- Quality flags: `9`

# Troubleshooting

## Common Errors

### "Platform name not found in lookup table"

**Problem**: Ship code not recognized

**Solution**: Update the platforms lookup table:
```{r update-platform}
# Check current platforms
con <- dbConnect(RSQLite::SQLite(), 'lookup.sqlite')
dbReadTable(con, "platforms")
dbDisconnect(con)

# Add missing platform (if needed)
# Contact database administrator or update LookupTables.R
```

### "EXPOCODE not properly generated"

**Problem**: Date format in BCD data is incorrect

**Solution**: Ensure dates are in `%m/%d/%Y` format (e.g., 05/14/2025)

### "Multiple DISTINCT units found"

**Problem**: Conflicting units for a parameter in lookup table

**Solution**: This indicates a data quality issue. Review the specific parameter and resolve the unit conflict before proceeding.

### "No sounding data found in BioChem"

**Problem**: Cannot retrieve bottom depth for stations

**Solution**: 
1. Verify mission descriptor is correct
2. Check if mission exists in BioChem
3. Ensure your BioChem credentials have proper access

## Validation Warnings

The conversion process includes automatic validation. Pay attention to warnings:

- **Latitude/Longitude out of range**: Check for sign errors or missing decimals
- **Parameter outside typical range**: May indicate data entry errors
- **Missing decimal places**: Precision loss in coordinates

# Next Steps

After generating your submission files:

1. **Review the output**: Open in Excel or R and spot-check values
2. **Check validation reports**: Address any warnings
3. **Prepare metadata**: Each platform requires additional metadata files
4. **Submit**: Follow platform-specific submission procedures (see submission guide vignette)

# Additional Resources

- **Submission Guide**: See `vignette("submission-guide")` for platform-specific instructions
- **BioChem Query**: `inst/BCD_QUERY.sql` for the standard extraction query
- **Lookup Tables**: `LookupTables.R` for managing reference data
- **Function Documentation**: `?convert_OCADS`, `?convert_CCHDO`, `?convert_SDG`

# Getting Help

Common issues and solutions:

1. **Update lookup tables**: If new ships or methods are added
2. **Check BioChem connection**: Verify credentials and VPN status
3. **Review data format**: Ensure BCD format matches expectations
4. **Contact**: Previous data manager or BioChem administrator

---

*Last updated: `r Sys.Date()`*
