---
title: "Platform Submission Guide: OCADS, CCHDO, and SDG"
author: "BIO Data Management Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Platform Submission Guide: OCADS, CCHDO, and SDG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This guide provides detailed instructions for submitting oceanographic data to three international data platforms:

1. **OCADS** - Ocean Carbon Data System
2. **CCHDO** - CLIVAR and Carbon Hydrographic Data Office
3. **SDG** - Sustainable Development Goals (UN)

Each platform has specific requirements, submission procedures, and metadata needs.

---

# OCADS: Ocean Carbon Data System

## About OCADS

OCADS is hosted by the Japan Agency for Marine-Earth Science and Technology (JAMSTEC) and serves as a repository for ocean carbon and related oceanographic data.

**Website**: https://oceans.imas.utas.edu.au/OCADS/ (or current URL)

## Data Requirements

### File Format
- **Format**: CSV (comma-separated values)
- **Encoding**: UTF-8
- **Quote style**: Standard (quotes allowed)
- **Missing data**: -999 for values, 9 for flags

### Required Parameters
At minimum, OCADS submissions should include:
- Station information (EXPOCODE, STNNBR, CASTNO)
- Position (LATITUDE, LONGITUDE)
- Date and time
- Depth/Pressure (DEPTH or CTDPRS)
- At least one measured parameter with quality flags

### Recommended Parameters for Carbon Studies
- **Nutrients**: NO2+NO3, PHSPHT, SILCAT
- **Carbonate system**: TCARBN, ALKALI, PH_TOT, PCO2
- **Physical**: CTDSAL, CTDTMP, CTDOXY
- **Biological**: CHLORA

## Preparing Your Submission

```{r ocads-prep}
library(BIOsubmissions)
library(tidyverse)

# Source credentials
source("C:/users/ogradye/desktop/biochem_creds.R")

# Load and prepare data
data <- read_csv("mission_BCD.csv")

# Convert to OCADS format
ocads_data <- convert_OCADS(data, biochem.password, biochem.user)

# Save the file
write_csv(ocads_data, "MISSION_OCADS_submission.csv")
```

## Required Metadata

All data values and quality control flags are pulled directly from BioChem, ensuring traceability
and consistency. However, you still need to prepare a separate metadata file (text or Word document) with:

1. **Cruise Information**
   - Cruise ID / Mission descriptor
   - Ship name
   - Dates (start and end)
   - Geographic region

2. **Principal Investigator**
   - Name
   - Institution
   - Email
   - ORCID (if available)

3. **Methods** (obtained from BioChem metadata and lab SOPs)
   - Sample collection procedures
   - Analytical methods for each parameter
   - Detection limits
   - Calibration procedures
   - Reference materials used (e.g., CRM batch numbers from BioChem)

4. **Quality Control** (documented in BioChem)
   - Replicate analysis frequency (handled automatically by package)
   - Accuracy estimates (from CRM analyses stored in BioChem)
   - Precision estimates (from replicate analyses in BioChem)
   - Known issues or limitations

5. **Data Processing**
   - Software used: BIOsubmissions R package
   - Conversion factors applied (automated by package)
   - Flag meanings: WOCE standards (BioChem 0-9 → WOCE 2,3,4,6,9)
   - Data source: BioChem database (extracted via BCD query)

## Submission Process

1. **Validate your data file**
   - Check that all required columns are present
   - Verify flag columns exist for each data column
   - Confirm date formats are correct (YYYY-MM-DD)
   - Review warnings from `convert_OCADS()`

2. **Prepare metadata document**
   - Use the template above
   - Be thorough - metadata is critical for data usability

3. **Submit via email or web portal**
   - Contact: (check OCADS website for current submission contact)
   - Subject line: "Data Submission - [MISSION_DESCRIPTOR]"
   - Attach data file and metadata
   - Request confirmation of receipt

4. **Respond to reviewer feedback**
   - OCADS may request clarifications or corrections
   - Common issues: coordinate precision, unit clarification, flag documentation

## Common OCADS Rejection Reasons

Based on historical submissions:

- ❌ **Latitude/Longitude precision loss**: Missing decimal places
- ❌ **Incorrect sign on longitude**: Western longitudes should be negative
- ❌ **Out-of-range values**: Coordinates outside expected region
- ❌ **Insufficient metadata**: Missing method descriptions
- ❌ **Unexplained flags**: Quality flags without documentation

---

# CCHDO: CLIVAR and Carbon Hydrographic Data Office

## About CCHDO

CCHDO is hosted by UC San Diego and focuses on high-quality hydrographic data, particularly from repeat hydrography programs like GO-SHIP.

**Website**: https://cchdo.ucsd.edu/
**Submission portal**: https://cchdo.ucsd.edu/submit

## Data Requirements

### File Format
- **Format**: CSV (comma-separated values)
- **Encoding**: UTF-8
- **Quote style**: **IMPORTANT - use `quote = 'none'`**
- **Missing data**: -999 for values, 9 for flags
- **Date format**: YYYYMMDD (no delimiters!)

### Column Naming
CCHDO uses standardized parameter names:
- `CTDPRS` (pressure, dbar)
- `CTDTMP` (temperature, °C ITS-90)
- `CTDSAL` (salinity, PSS-78)
- `CTDOXY` (oxygen, µmol/kg)
- `NO2+NO3` (nitrate+nitrite, µmol/kg)
- `PHSPHT` (phosphate, µmol/kg)
- `SILCAT` (silicate, µmol/kg)

### Unit Row
**Critical**: The first row must contain units, not data!

```
EXPOCODE,STNNBR,DEPTH,CTDSAL,NO2+NO3,NO2+NO3_FLAG_W
18QL20230504,AR7W01,METERS,PSU,µmol/kg,
```

## Preparing Your Submission

```{r cchdo-prep}
library(BIOsubmissions)
library(tidyverse)

# Start from OCADS output
ocads_data <- convert_OCADS(data, biochem.password, biochem.user)

# Convert to CCHDO format
cchdo_data <- convert_CCHDO(ocads_data)

# CRITICAL: Use quote = 'none'
write_csv(cchdo_data, 
          "MISSION_CCHDO_submission.csv",
          quote = 'none')
```

## CCHDO Exchange Format Requirements

Key differences from OCADS:

1. **Date format**: YYYYMMDD (not YYYY-MM-DD)
2. **Column names**: LATITUDE, LONGITUDE (not BTL_LAT, BTL_LON)
3. **Station names**: Alphanumeric only (special characters removed)
4. **Quote style**: None (`quote = 'none'`)
5. **NH4 vs NH3**: CCHDO prefers NH4 nomenclature

## Required Metadata

CCHDO requires extensive documentation. All data values and QC flags come directly from BioChem:

1. **Cruise Summary**
   - EXPOCODE (ship code + date)
   - Chief Scientist
   - Ship name and country
   - Ports (departure and arrival)
   - Dates
   - Funding agency

2. **Parameters Measured**
   - For EACH parameter, provide:
     - PI/Analyst name and institution
     - Method reference (publication or SOP)
     - Instrument make/model
     - Calibration procedure
     - Accuracy and precision (from BioChem CRM analyses)
     - CRM batch numbers (stored in BioChem metadata)

3. **Bottle Data Processing**
   - Fire order and bottle IDs (from BioChem)
   - Sample collection notes
   - Known issues or suspect data
   - Data source: BioChem database with automated validation

4. **Supporting Files** (if available)
   - CTD data (if not already submitted)
   - Cruise report
   - Method documentation (SOPs)
   - Station map
   
**Note:** The BIOsubmissions package automatically extracts all data and QC flags from BioChem,
ensuring complete traceability. Metadata documentation should reference BioChem as the source
database and note any additional quality control applied during BioChem upload.

## Submission Process

### Online Submission Portal

1. **Create an account** at https://cchdo.ucsd.edu/submit
   - Use your institutional email
   - Save your login credentials

2. **Start a new submission**
   - Click "Submit Data"
   - Select "Bottle Data"
   - Enter EXPOCODE

3. **Upload data file**
   - Use the CCHDO-formatted CSV
   - Portal will perform automatic validation
   - Review and address any errors

4. **Complete metadata form**
   - Fill in all required fields
   - Upload supporting documents
   - Be thorough - incomplete submissions are delayed

5. **Submit for review**
   - CCHDO staff will review (usually 2-4 weeks)
   - Respond promptly to questions
   - Revisions may be requested

### Email Submission (Alternative)

If the portal is unavailable:

- **To**: cchdo@ucsd.edu
- **Subject**: Data Submission - [EXPOCODE]
- **Attachments**: Data file + metadata document
- **Body**: Brief description and contact information

## Common CCHDO Rejection Reasons

Based on historical submissions:

- ❌ **Quote marks in file**: Must use `quote = 'none'`
- ❌ **Wrong date format**: Must be YYYYMMDD
- ❌ **Missing unit row**: First row must contain units
- ❌ **Non-standard parameter names**: Must match CCHDO dictionary
- ❌ **Insufficient metadata**: Especially method documentation
- ❌ **Undocumented flags**: All non-2 flags should be explained

---

# SDG: Sustainable Development Goals

## About SDG

The UN Sustainable Development Goals include SDG 14 (Life Below Water) with specific ocean chemistry indicators. Data contributes to global assessments of ocean health.

**Website**: https://sdg.data.un.org/

## Data Requirements

SDG submissions typically follow OCADS format but with additional requirements:

### Mandatory Parameters for SDG 14.3.1
- **Carbonate system** (at least 2 of):
  - pH (PH_TOT)
  - Total alkalinity (ALKALI)
  - Dissolved inorganic carbon (TCARBN)
  - pCO2 (PCO2)

### Spatial/Temporal Coverage
- Data should contribute to spatial or temporal coverage gaps
- Coastal and shelf regions are priorities
- Repeat measurements are valuable for trend detection

## Preparing Your Submission

```{r sdg-prep}
library(BIOsubmissions)
library(tidyverse)

# SDG typically uses OCADS format
ocads_data <- convert_OCADS(data, biochem.password, biochem.user)

# Verify carbonate parameters are present
required_params <- c("PH_TOT", "ALKALI", "TCARBN", "PCO2")
available <- required_params[required_params %in% names(ocads_data)]

if (length(available) >= 2) {
  message("✓ Sufficient carbonate parameters for SDG submission")
} else {
  warning("⚠ SDG requires at least 2 carbonate parameters")
}

write_csv(ocads_data, "MISSION_SDG_submission.csv")
```

## Required Metadata

SDG submissions should emphasize:

1. **Policy Relevance**
   - How data contributes to SDG 14.3.1
   - Regional or national context
   - Monitoring program description

2. **Data Stewardship**
   - Long-term data management plan
   - Data sharing agreements
   - Update frequency

3. **Standard Metadata**
   - All elements required for OCADS (see above)

## Submission Process

SDG data is typically submitted through:

1. **OCADS** (primary route for ocean chemistry)
2. **GOOS/IODE** - Integrated Ocean Observing System
3. **National data centers** with SDG reporting

**Recommended**: Submit to OCADS and indicate SDG relevance in metadata.

---

# Data Quality Checklist

Before submitting to ANY platform, verify:

## ✅ File Format
- [ ] CSV format with proper encoding
- [ ] Correct delimiter (comma)
- [ ] Quote style matches platform requirements
- [ ] Unit row present (first row for CCHDO/OCADS)

## ✅ Data Completeness
- [ ] All required columns present
- [ ] Flag column for each data column
- [ ] No completely empty columns
- [ ] Station/cast/sample identifiers unique

## ✅ Data Quality
- [ ] Latitude range: -90 to 90
- [ ] Longitude range: -180 to 180  
- [ ] Depth values positive
- [ ] Dates in correct format
- [ ] No NA values (use -999 instead)
- [ ] Parameter values within oceanographic ranges

## ✅ Metadata
- [ ] PI/analyst contact information
- [ ] Method descriptions for each parameter
- [ ] Calibration and QC procedures
- [ ] Known issues documented
- [ ] Cruise/mission summary

## ✅ Documentation
- [ ] Flag meanings explained
- [ ] Units clearly specified
- [ ] Conversions documented
- [ ] References cited

---

# After Submission

## Timeline Expectations

- **OCADS**: 2-4 weeks for initial review
- **CCHDO**: 2-6 weeks depending on complexity
- **SDG**: Variable (depends on routing)

## Following Up

If you haven't heard back in the expected timeline:

1. Check spam folder for emails
2. Email the submission contact
3. Reference your submission ID or EXPOCODE
4. Be patient but persistent

## Data Citations

Once accepted:

- OCADS: DOI assigned for dataset
- CCHDO: Data available at https://cchdo.ucsd.edu/cruise/[EXPOCODE]
- SDG: Included in global assessments

Request proper citations in publications using the data!

---

# Contact Information

## BIO Data Management
- **Current contact**: [Update with current data manager]
- **BioChem support**: biochem@dfo-mpo.gc.ca

## Platform Contacts
- **OCADS**: Check https://oceans.imas.utas.edu.au/OCADS/ for current contact
- **CCHDO**: cchdo@ucsd.edu
- **SDG/UN**: https://unstats.un.org/sdgs/

---

# Appendix: Historical Issues Log

Common problems encountered in past submissions:

## Issue: Missing Negative on Longitude
**Symptom**: Western Atlantic longitudes appear as positive
**Cause**: Data entry or export error from BioChem
**Solution**: Verify and manually correct if needed

## Issue: Latitude Missing Decimals  
**Symptom**: All latitudes are whole numbers (e.g., 44 instead of 44.267)
**Cause**: Excel formatting or CSV export issue
**Solution**: Re-export from source with full precision

## Issue: Depth Units Confusion
**Symptom**: CCHDO rejects because DEPTH is in centimeters not meters
**Cause**: BioChem stores depth in cm
**Solution**: `convert_OCADS()` handles this automatically

## Issue: Multiple Unit Conflicts
**Symptom**: Error during conversion about multiple units
**Cause**: Lookup table has conflicting entries
**Solution**: Review and update lookup table to resolve

---

*Last updated: `r Sys.Date()`*
*For questions about this guide, contact the BIO data management team.*
